---
date: "2022-07-09T00:00:00Z"
Summary: The Capstone Project for the Google Data Analytics Specialization
Title: Google Data Analytics Capstone
---

# Google Data Analytics Capstone: Analysis Of 2019 Cyclistic Bike Share

## Prompt

"You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations."

*Understand how annual members and casual riders differ
*Identify trends to know why casual riders would buy annual membership
*To use marketing strategies to convert casual riders to annual members.

```{r}
# Loading needed R Packages

library(tidyverse) #for analyzing data
library(lubridate) #to work with dates
library(ggplot2) #to create graphs
```

## STEP 1: COLLECT DATA

```{r}
#Download/import data

q1_2019 <- read.csv("Divvy_Trips_2019_Q1.csv")
q2_2019 <- read.csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read.csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read.csv("Divvy_Trips_2019_Q4.csv")
```

## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

```{r}
# Checking and previewing file names

colnames(q1_2019)
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
```


```{r}
fixed_q2_2019 <- rename(q2_2019, trip_id = "X01...Rental.Details.Rental.ID", 
                      start_time = "X01...Rental.Details.Local.Start.Time",
                      end_time = "X01...Rental.Details.Local.End.Time",
                      bikeid = "X01...Rental.Details.Bike.ID",
                      tripduration = "X01...Rental.Details.Duration.In.Seconds.Uncapped",
                      from_station_id = "X03...Rental.Start.Station.ID",
                      from_station_name = "X03...Rental.Start.Station.Name",
                      to_station_id = "X02...Rental.End.Station.ID",
                      to_station_name = "X02...Rental.End.Station.Name",
                      usertype = "User.Type",
                      gender = "Member.Gender",
                      birthyear = "X05...Member.Details.Member.Birthday.Year")

```


```{r}
# Checking and previewing file names after fixing

colnames(q1_2019)
colnames(fixed_q2_2019)
colnames(q3_2019)
colnames(q4_2019)
```

```{r}
# Checking the dataframes

str(q1_2019)
str(fixed_q2_2019)
str(q3_2019)
str(q4_2019)
```

```{r}
# Fixing dataframes

q4_2019 <-  mutate(q4_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
q3_2019 <-  mutate(q3_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
fixed_q2_2019 <-  mutate(fixed_q2_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
q1_2019 <-  mutate(q1_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
```


```{r}
# Combining into one big dataframe

all_trips <- bind_rows(q1_2019, fixed_q2_2019, q3_2019, q4_2019)
```


```{r}
# Removing unneeded data for this analysis.

all_trips <- all_trips %>%  
  select(-c(birthyear, gender))
```

## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

```{r}
# Inspecting before cleaning

colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

```{r}
# In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"

(all_trips <- rename(all_trips
                   ,member_casual = usertype))

all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))
```

```{r}
# Checking to make sure the proper number of observations were reassigned

table(all_trips$member_casual)
```


```{r}
# Adding columns that list the date, month, day, and year of each ride

all_trips$date <- as.Date(all_trips$start_time) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

```{r}
# Adding a "ride_length" calculation to all_trips (in seconds)

all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)
```

```{r}
# Inspecting the structure of the columns

str(all_trips)
```

```{r}
# Convert "ride_length" from Factor to numeric so we can run calculations on the data

is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

```

```{r}
# Removing "bad" data
# The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

all_trips_v2 <- all_trips[!(all_trips$from_station_name == "HQ QR" | all_trips$ride_length<0),]
```

## STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

```{r}
# Descriptive analysis on ride_length (all figures in seconds)

mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```

```{r}
summary(all_trips_v2$ride_length)
```

```{r}
# Comparing members and casual users

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```


```{r}
# Averaging ride time by each day for members vs casual users

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r}
# Fixing out of order days.

all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


```{r}
# Averaging ride time by each day for members vs casual users

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r}
# Analyzing ridership data by type and weekday

all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()                            #calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>%         # calculates the average duration
  arrange(member_casual, weekday)                                # sorts
```


### Visualizing Data

```{r}
# Visualizing the number of rides by rider type

all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x = "Weekday", y ="Number of Rides",fill = "Member or Casual",title ="Number of Rides in a Day") 
```

```{r}
# Visualizing for average duration

all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x = "Weekday", y ="Trip Duration(seconds)",fill = "Member or Casual",title ="Average Duration per Ride") 
```

```{r}
# Month-by-month data

all_trips_v2 %>%
  group_by(month, member_casual) %>%
  summarize(total = n()) %>%
  arrange(month, member_casual) %>%
  ggplot(aes(x = month, y = total, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x = "Months", y ="Total Rides",fill = "Member or Casual",title ="Total Rides per Month")
```

```{r}
# Hour-by-hour data

all_trips_v2$start_hour <- as.POSIXct(all_trips_v2$start_time, "%Y-%m-%d %H:%M:%S")
```

```{r}
# Visualized hour-by-hour data

all_trips_v2 %>%
  group_by(hour_of_day = hour(round_date(start_hour, 'hour'))) %>% 
  group_by(hour_of_day, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  arrange(-number_of_rides) %>% 
  ggplot(aes(x = hour_of_day, y = number_of_rides, fill = member_casual)) +
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23)) +
  labs(x = "Time of the Day", y = "Number of Rides", 
       fill = "Member or Casual",
       title = "Number of Rides by Hour")
```


## STEP 5: FURTHER ANALYSIS

### Findings:

* Casual riders spend more time, on average, on the bike service than member riders in a day.
* However, member riders use the bikes more on a daily basis (number of daily rides).
* The number of daily rides for casual riders gets closer to member riders on weekend.
* Month-by-month data shows that it peaks in the summer, and falls off in the other seasons.
* Hour-by-hour data shows that there are different trends for each group.
* Member riders peak in the morning and afternoon.
* Casual riders keep consistent from noon to night.

### End of Capstone!
